---
title: "Introduction to `ggplot2`"
author: "Carla Giner Delgado"
date: "January 12, 2018"
output:
  html_document:
    toc: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.show = "hold")
```

Overview of the basics and underlying structure of the `R` graphical package `ggplot2` [@Wickham2010].

This document is written in an `R Markdown` and allows `R` code chunks to be interleaved with `Markdown` formatted text (more info about [`R Markdown`](http://rmarkdown.rstudio.com), and about [`Markdown`](http://rmarkdown.rstudio.com/authoring_pandoc_markdown.htm)).



Components of `ggplot2` 
=======================

The syntax of `ggplot2` is based on _The Grammar of Graphics_ [@Wilkinson2010], and each plot can be obtained by a series of _specifications_, that are given as _layers_. Let's have a quick look to the different parts. [Full reference](http://ggplot2.tidyverse.org/reference/index.html).

```{r load_package}
# install.packages("ggplot2") # If needed, install ggplot2
library(ggplot2) # Load the package
```


Tidy data set
-------------

First of all, `ggplot2` expects a tidy data frame as input. The format needs to be in a long layout (opposed to a wide layout). Long data has one observation per row and one variable per column:

Student | Programming language | Knowledge
--------|----------------------|-----------
John    | R                    | 3 High
John    | C                    | 1 Low
Mary    | R                    | 2 Medium
Mary    | C                    | 3 High


The same information in a wide layout (not suitable for `ggplot2`) would be:

Student | R        | C        
--------|----------|-------
John    | 3 High   | 1 Low
Mary    | 2 Medium | 3 High


We will use a simple data frame with these data to test `ggplot2` commands. Later, we will use real-life (although tidy) data sets in the [examples](#expression).

```{r simple_data}
programming_level <- data.frame(
    student = c("John", "John", "Mary", "Mary"),
    programming_language = c("R", "C", "R", "C"),
    knowledge = c(3, 1, 2, 3)
    )
```

Plot calls start with `ggplot(data = <data>)` and then extra specifications are added using the sign `+`.


Geoms and aesthetic mapping to variables
----------------------------------------

The main layers tell `ggplot` what variables from the data to represent and how are they going to be represented.

We typically add a layer with a `geom_*()` function. There are geoms (geometric objects) to represent common graphs such as points, lines, bars, polygons or text. Each geom has a set of aesthetic attributes (x position, y position, colour, shape...) that can be mapped to variables in our data through the function parameter `mapping`, using the function `aes()`.

```{r geom_examples, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language))

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = knowledge))
```

Aesthetics can also be modified for all observations by calling them outside of the `aes()` function.

```{r constant_aes, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language), alpha = 0.5, size = 5)

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = knowledge), colour = "white", size = 5)
```

Type `?geom_*` (function help) in order to know what aesthetics the geom can take.


Stats and position adjustment
-----------------------------

Statistical transformations of the variables (e.g. represent counts of each value) and the adjustment of the position can also be specified in the geom function with the parameters `stat` and `position`. Alternatively we can use `geom_*()` functions for common representations that have already the required transformations and position adjustments as defaults.

For example, to make a histogram we can change the statistical transformation and the position of `geom_bar()` or we can use the function `geom_histogram()`, that has the right defaults for `stat` and `position`.

```{r stat_position, fig.height = 3, fig.width = 4, message = FALSE}
# geom_bar
ggplot(data = programming_level) +
    geom_bar(aes(x = knowledge), stat = "bin", position = "stack")

# genom_histogram
ggplot(data = programming_level) +
    geom_histogram(aes(x = knowledge))
```

Basic position values:  

* identity: Don't adjust position
* stack: Stack overlapping objects on top of each another
* dodge: Dodge overlapping objects side-to-side
* nudge: Nudge points a fixed distance
* jitter: Jitter points to avoid overplotting


Scales
------

The values taken by the aesthetics mapped to variables can be controlled by another set of functions called `scale_<aesthetic>_<name>`. [All available scales](http://ggplot2.tidyverse.org/reference/index.html#section-scales)

```{r scales, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language)) + 
    scale_shape_manual(values = c(3, 8))+
    scale_y_continuous(breaks = c(1, 2, 3), labels = c("Low", "Medium", "High"), name = "Programming level")

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = as.factor(knowledge))) +
    scale_fill_brewer(type = "seq", palette = "Greens", name  = "Level", labels = c("Low", "Medium", "High"))

```

There are some helper functions for common tasks.

* `labs()`: set axes titles, plot title, subtitle...
* `lims()`: set axis limits

```{r labs, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language)) + 
    scale_shape_manual(values = c(3, 8))+
    scale_y_continuous(breaks = c(1, 2, 3), labels = c("Low", "Medium", "High"), name = "Programming level") +
    labs(shape = "Language", title = "Programming knowledge", x = "Student name")

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = as.factor(knowledge))) +
    scale_fill_brewer(type = "seq", palette = "Greens", name  = "Level", labels = c("Low", "Medium", "High")) +
    labs(y = "Programming language", title = "Programming knowledge")
```


Coordinates
-----------

There are two coordinates systems available:

* Cartesian, the default. Some `coord_*()` functions can modify the system slightly (e.g. swap axes with `coord_flip()`).
* Polar, using the function `coord_polar()`.

```{r polar, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language)) + 
    scale_shape_manual(values = c(3, 8))+
    scale_y_continuous(breaks = c(1, 2, 3), labels = c("Low", "Medium", "High"), name = "Programming level") +
    labs(shape = "Language", title = "Programming knowledge", x = "Student name") +
    coord_polar()

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = as.factor(knowledge))) +
    scale_fill_brewer(type = "seq", palette = "Greens", name  = "Level", labels = c("Low", "Medium", "High")) +
    labs(y = "Programming language", title = "Programming knowledge") +
    coord_polar()
```



Facets
------

We can also create a set of subplots using variables from the data. It is especially useful when representing many variables. There are two functions available `facet_grid()` and `facet_wrap()`.

```{r facets, fig.height = 3, fig.width = 4}
ggplot(data = programming_level) +
    geom_bar(mapping = aes(x = student, y = knowledge), stat = "identity") + 
    scale_y_continuous(breaks = c(1, 2, 3), labels = c("Low", "Medium", "High"), name = "Programming level") +
    labs(title = "Programming knowledge", x = "Student name") +
    facet_grid(.~programming_language)
```



Themes
------

Themes control how the elements of the plot are represented (excluding the data themselves). The default theme can be substituted by another (e.g. `theme_classic()`) or individual elements of the theme can be modified with the function `theme()` (`?theme` for the list of plot elements).

```{r themes, fig.height = 3, fig.width = 4}
# Scatterplot
ggplot(data = programming_level) +
    geom_point(mapping = aes(x = student, y = knowledge, shape = programming_language)) + 
    scale_shape_manual(values = c(3, 8))+
    scale_y_continuous(breaks = c(1, 2, 3), labels = c("Low", "Medium", "High"), name = "Programming level") +
    labs(shape = "Language", title = "Programming knowledge", x = "Student name") +
    theme_classic(base_size = 16)

# Heatmap
ggplot(data = programming_level) +
    geom_tile(mapping = aes(x = student, y = programming_language, fill = as.factor(knowledge))) +
    scale_fill_brewer(type = "seq", palette = "Greens", name  = "Level", labels = c("Low", "Medium", "High")) +
    labs(y = "Programming language", title = "Programming knowledge") + 
    theme(legend.position = "left")
```


Real-life example: Gene expression {#expression}
==================================

Data set
--------

[GTEx](https://gtexportal.org/home/) median gene expression values in TMP for 8 tissues and 54225 genes.

```{r read_data}
# Median expression values
expression_data<-read.table(file = "data/expression_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
expression_data$Expressed <- expression_data$median_expression > 0

ggplot(expression_data) + 
    geom_line(aes(tissue, median_expression, group = gene_id), alpha = 0.01)+
    # geom_boxplot(aes(tissue, median_expression, fill = gene_type ))+
    scale_y_log10()+
    facet_grid(gene_type~gene_annotation)


ggplot(expression_data) + 
    # geom_line(aes(tissue, median_expression, group = gene_id), alpha = 0.001)+
    geom_boxplot(aes(gene_type, median_expression, fill = tissue))+
    scale_y_log10()#+
    #facet_grid(gene_type~.)

```


Types of genes
--------------
- Which kind of genes did each annotation source annotate?
- Which are more expressed (median)


Expressed genes
---------------
- Which tissue has the greatest number of genes expressed?



Constitutive genes
------------------
- What genes are expressed at >100 TPM?





References
==========
